---
title: "PCA myeloma multiple"
author: "Esther Tercero Atencia"
date: "2023-05-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plotly)
```

## In this document we will load the MYLEOMA MULTIPLE dataset already filtered and plot Principal Component Analysis before and after normalization and correction by possible confounding variables. 

## Some details about the procedure:

> Normalization is performed with vsn approach (code for TMM normalization is also avaiable)
> PCA helps to evaluate any possible effect of the intrinsic properties of the sample (code for 2d and 3d PCA avaiable)
> In MYELOMA MULTIPLE dataset we've got as confounding factors: group, age, gender, rna batch and library batch. For Age (continuous variable) we created a new discrete variable called `age2`, dividing samples in four groups "<25 years", "25-40 years","45-65 years", ">65 years". RNA and library showed collinearity, so here we'll provide the correction using both alone. Then the best PCA (the one that shows more correction) helped us to finally select the library batch correction.
> Some races (other, asian) were only represented by one sample, so batch couldn't be corrected. Those samples were discarded because preliminar PCAs separate samples by race. 

```{r}
# metadata
metadata_mm <- read.csv(here("data/metadata/all_metadata_prep.csv")) %>% 
  filter(outcome == "MM" | outcome == "HD") %>%
  filter(grepl("PP", sample)) %>%
  dplyr::select(-X)

# filtered counts
filt_mm <- read.csv(here("data/filtered_data/filtered_dedupCounts_mm.csv")) %>%
  column_to_rownames(var = "GeneID") %>%
  dplyr::select(metadata_mm$sample)


group <- metadata_mm$group
age <- metadata_mm$age
age2 <-  cut(metadata_mm$age, 
             breaks = c(0, 25, 45, 65, 100), 
             labels = c("<25", "25-40","45-65", ">65"))
gender <- metadata_mm$gender
race <- metadata_mm$race
rna_batch <- metadata_mm$RNA_extraction
lib_batch <- metadata_mm$Library_preparation
```

```{r}
## pca without normalization and correction
pca <- prcomp(t(filt_mm))
pcamat1 <- as.data.frame(pca$x) %>%
  rownames_to_column(var="sample") %>%
  merge(metadata_mm, by="sample")

var_exp <- round(pca$sdev^2/sum(pca$sdev^2)*100,2)

## split by group
p1 <- ggplot(pcamat1,aes(PC1,PC2,colour=group))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "group")
  ) +
    ggsci::scale_color_npg()

## split by age
p2 <- ggplot(pcamat1,aes(PC1,PC2,colour=age2))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "age")
  ) +
    ggsci::scale_color_npg()

## split by gender
p3 <- ggplot(pcamat1,aes(PC1,PC2,colour=gender))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "gender")
  ) +
    ggsci::scale_color_npg()

## split by race
p4 <- ggplot(pcamat1,aes(PC1,PC2,colour=race))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "race")
  ) +
    ggsci::scale_color_npg()

## split by rna_batch
p5 <- ggplot(pcamat1,aes(PC1,PC2,colour=RNA_extraction))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "rna batch")
  ) +
    ggsci::scale_color_npg() +
  scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9", "#8B4513", "#F1C40F", "#2E86C1", 
                              "#27AE60", "#8E44AD", "#C0392B", "#000000","#16c000",  "#3498DB"))

## split by lib_batch
p6 <- ggplot(pcamat1,aes(PC1,PC2,colour=Library_preparation))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "lib batch")
  ) +
    ggsci::scale_color_npg()

pca_2d_mm <- ggarrange(p1,p2,p3,p4,p5,p6, nrow = 2, ncol =3)
pca_3d_mm <- plot_ly(pcamat1, x = ~PC1, y = ~PC2, z = ~PC3, color = ~group, colors = c('#636EFA','#EF553B') ) %>%
  add_markers()

svg(here("/results/PCA/PCA_2d_mm.svg"), width =20, height = 10); pca_2d_mm; # dev.off()
```

```{r}
## pca with normalization and correction
## calculate norm factors for TMM normalization
norm.factors <- DGEList(counts = filt_mm, samples = metadata_mm, group = metadata_mm$outcome)
norm.factors <- calcNormFactors(norm.factors, method = "TMM") 

## design matrix for correction
design = model.matrix(~group + age + gender) 

## normalization with vsn
norm_counts_mm_vsn <- limma::removeBatchEffect(limma::normalizeVSN(norm.factors), 
                                             batch = lib_batch, 
                                             design = design) %>% 
   data.frame()

## normalization with tmm
norm_counts_mm_tmm <- limma::removeBatchEffect(cpm(norm.factors, log = TRUE), 
                                             batch = lib_batch, 
                                             design = design) %>% 
   data.frame()


## change norm_mm WITH THE SPECIFYC NORMALIZATION to continue

norm_mm <- norm_counts_mm_vsn
#norm_mm <- norm_counts_mm_tmm

pca <- prcomp(t(norm_mm))
pcamat1 <- as.data.frame(pca$x) %>%
  rownames_to_column(var="sample") %>%
  cbind(metadata_mm %>% dplyr::select(-sample))

var_exp <- round(pca$sdev^2/sum(pca$sdev^2)*100,2)

## split by group
p1 <- ggplot(pcamat1,aes(PC1,PC2,colour=group))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "group")
  ) +
    ggsci::scale_color_npg()

## split by age

p2 <- ggplot(pcamat1,aes(PC1,PC2,colour=age2))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "age")
  ) +
    ggsci::scale_color_npg()

## split by gender
p3 <- ggplot(pcamat1,aes(PC1,PC2,colour=gender))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "gender")
  ) +
    ggsci::scale_color_npg()

## split by race
p4 <- ggplot(pcamat1,aes(PC1,PC2,colour=race))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "race")
  ) +
    ggsci::scale_color_npg()

## split by rna_batch
p5 <- ggplot(pcamat1,aes(PC1,PC2,colour=RNA_extraction))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "rna batch")
  ) +
    ggsci::scale_color_npg() +
  scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9", "#8B4513", "#F1C40F", "#2E86C1", 
                              "#27AE60", "#8E44AD", "#C0392B", "#000000","#16c000",  "#3498DB"))

## split by lib_batch
p6 <- ggplot(pcamat1,aes(PC1,PC2,colour=Library_preparation))+
  geom_point(size=2)+
  theme_bw() +
  labs(
    x = paste("PC1 (", var_exp[1],"%)", sep=""),
    y = paste("PC2 (", var_exp[2],"%)", sep=""),
    title = paste0("PCA splitted by ", "lib batch")
  ) +
    ggsci::scale_color_npg()

pca_2d_mm <- ggarrange(p1,p2,p3,p4,p5,p6, nrow = 2, ncol =3)
pca_3d_mm <- plot_ly(pcamat1, x = ~PC1, y = ~PC2, z = ~PC3, color = ~group, colors = c('#636EFA','#EF553B') ) %>%
  add_markers()

svg(here("results/PCA/PCA_2d_mm_norm_vsn.svg"), width =20, height = 10); pca_2d_mm ; dev.off()
#svg(here("results/PCA/PCA_2d_mm_norm_tmm.svg"), width =20, height = 10); pca_2d_mm ; dev.off()


```

