---
title: "Normalization and batch correction of the datasets"
author: "Esther Tercero Atencia"
date: "2023-04-25"
output: pdf_document
---
In this document we will normalize (vsn) and correct batch effects of the MYELOMA MULTIPLE and SIMULATED count matrices. The code for filter the HEPATOCELLULAR dataset is also included, but muted because it is not the main of the work.

Here also we will split the datasets following the division of the reference dataset or with a randomize approach.

Normalization + batch removal is muted because next steps (DiffExpression Analyses) have an step of batch correction in the pipeline of analyses, but the code is shared here in case it is of interest.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(edgeR)
library(DESeq2)
library(tidymodels)
library(here)
source(here("utils/functions.R"))
```

### Metadata and filtered counts loading
```{r}
## metadata
metadata_all <- read.csv("../data/metadata/all_metadata_prep.csv") %>%
  apply(.,2,as.character) %>%
  as.data.frame()
```


### Filtered counts loading and discovery/validation split
```{r}
#------- MM
## Sample division for Myeloma multiple is performed as reported in the reference
## publication (Roskams-Hieter et al.). 

# metadata discovery
metadata_mm_discovery <- metadata_all %>%
  filter(outcome == "HD" | outcome == "MM") %>%
  filter(grepl("PP",sample)) %>%
  mutate(outcome = as.factor(outcome), 
         age = as.numeric(age), 
         gender = as.factor(gender),
         RNA_extraction = as.factor(RNA_extraction),
         Library_preparation = as.factor(Library_preparation), 
         set = as.factor(set), 
         race = as.factor(race),
         group = as.factor(group),
         age_fact = as.factor(cut(age, breaks=4))) 
write_csv(metadata_mm_discovery, file = here("data/metadata/metadata_mm_discovery.csv"))

# metadata_validation
metadata_mm_validation <- metadata_all %>%
  filter(outcome == "HD" | outcome == "MM") %>%
  filter(grepl("RSQ",sample)) %>%
  mutate(outcome = as.factor(outcome), 
         age = as.numeric(age), 
         gender = as.factor(gender),
         RNA_extraction = as.factor(RNA_extraction),
         Library_preparation = as.factor(Library_preparation), 
         set = as.factor(set), 
         race = as.factor(race),
         group = as.factor(group),
         age_fact = as.factor(cut(age, breaks=4))) 
write_csv(metadata_mm_validation, file = here("data/metadata/metadata_mm_validation.csv"))

# filtered counts
filtered_counts_mm <- read.csv(here("data/filtered_data/filtered_dedupCounts_mm.csv")) %>% 
  column_to_rownames("GeneID")
## discovery split
filtered_counts_mm_discovery <- filtered_counts_mm %>%
  select(grep("PP",colnames(filtered_counts_mm)))
write_csv(filtered_counts_mm_discovery %>% rownames_to_column(var = "GeneID"), here("data/filtered_data/filtered_mm_discovery.csv"))
## validation split
filtered_counts_mm_validation <- filtered_counts_mm %>%
  select(grep("RSQ",colnames(filtered_counts_mm)))
write_csv(filtered_counts_mm_validation %>% rownames_to_column(var = "GeneID"), here("data/filtered_data/filtered_mm_validation.csv"))
```

```{r}
#------- HC
## Sample division for Hepatocellular cancer is considered awkwardly divided in the reference
## publication (Roskam-Hieter et al.). [See Thesis theory]
## Resampling is needed here to get a balance and proper discovery/validation division
# set.seed(20)
# filtered_counts_hc <- read.csv(here("data/filtered_data/filtered_dedupCounts_hc.csv")) %>% 
#   column_to_rownames("GeneID")
# 
# hcc_data <- cbind(outcome = metadata_all 
#                   %>% filter(outcome == "HD" | outcome == "HC") 
#                   %>% dplyr::select(outcome),
#                   filtered_counts_hc 
#                   %>% t() 
#                   %>% data.frame())
# 
# hcc_split <- rsample::initial_split(hcc_data,            # Data to split
#                                      prop = 0.8,         # Proportion of data for training set
#                                      strata = outcome)   # Stratification variable
# filtered_counts_hc_discovery <- training(hcc_split) %>% 
#   rownames_to_column(var = "SampleID") %>%
#   dplyr::select(-outcome)
# write_csv(filtered_counts_hc_discovery, here("data/filtered_data/filtered_hc_discovery.csv"))
# 
# filtered_counts_hc_validation <- testing(hcc_split) %>%
#   rownames_to_column(var = "SampleID") %>%
#   dplyr::select(-outcome)
# write_csv(filtered_counts_hc_validation, here("data/filtered_data/filtered_hc_validation.csv"))
# 
# metadata_hc_discovery <- metadata_all %>% 
#   filter(metadata_all$sample %in% filtered_counts_hc_discovery$SampleID) %>% 
#   mutate(outcome = as.factor(outcome), 
#          age = as.numeric(age),
#          gender = as.factor(gender),
#          RNA_extraction = as.factor(RNA_extraction),
#          Library_preparation = as.factor(Library_preparation),
#          set = as.factor(set),
#          race = as.factor(race),
#          group = as.factor(group),
#          age_fact = as.factor(cut(age, breaks=4))) 
# write_csv(metadata_hc_discovery, here("data/metadata/metadata_hc_discovery.csv"))
# 
# metadata_hc_validation <- metadata_all %>% 
#   filter(metadata_all$sample %in% filtered_counts_hc_validation$SampleID) %>%
#   mutate(outcome = as.factor(outcome), 
#          age = as.numeric(age),
#          gender = as.factor(gender),
#          RNA_extraction = as.factor(RNA_extraction),
#          Library_preparation = as.factor(Library_preparation),
#          set = as.factor(set),
#          race = as.factor(race),
#          group = as.factor(group),
#          age_fact = as.factor(cut(age, breaks=4))) 
# write_csv(metadata_hc_validation, here("data/metadata/metadata_hc_validation.csv"))
```

```{r}
#------- Simulated data
# filtered counts
sims_filtered <- read.csv(here("data/filtered_data/filtered_dedupCounts_sim_rep.csv")) %>% 
  column_to_rownames(var = "GeneID") %>% 
  t() %>% 
  data.frame() %>% 
  rownames_to_column(var = "SampleID")

# metadata s
metadata_sims <- utils::read.csv(here("data/simulated_data/metadata_rep")) %>% 
  mutate(SampleID = GeneID, .keep = "unused")

all_set <- merge(metadata_sims, sims_filtered, by = "SampleID") %>% 
  column_to_rownames(var = "SampleID") 
cell_split <- initial_split(all_set,  
                            strata = group)

sim_discovery <- training(cell_split)
sim_validation  <- testing(cell_split)

metadata_discovery <- cbind(rownames(sim_discovery), sim_discovery$group) %>% 
  data.frame()
colnames(metadata_discovery) <- c("sampleID", "outcome")

metadata_validation <- cbind(rownames(sim_validation), sim_validation$group)  %>% 
  data.frame()
colnames(metadata_validation) <- c("sampleID", "outcome")

sim_discovery <- sim_discovery %>% 
  dplyr::select(-group) %>% 
  t() %>% 
  data.frame()
write_csv(sim_discovery %>% rownames_to_column(var = "GeneID"), here("data/filtered_data/filtered_sim_rep_discovery.csv"))

sim_validation <- sim_validation %>% 
  dplyr::select(-group) %>% 
  t() %>% 
  data.frame()
write_csv(sim_validation %>% rownames_to_column(var = "GeneID"), here("data/filtered_data/filtered_sim_rep_validation.csv"))
```



### Normalized and batch-corrected count matrix aplplying VSN normalization
```{r}
#------- MM
## discovery test
norm.factors <- DGEList(counts = filtered_counts_mm_discovery, samples = metadata_mm_discovery, group = metadata_mm_discovery$outcome)
norm.factors <- calcNormFactors(norm.factors, method = "TMM")

group <- metadata_mm_discovery$group
age <- metadata_mm_discovery$age
age_fact <- metadata_mm_discovery$age_fact
gender <- metadata_mm_discovery$gender
lib_batch <- metadata_mm_discovery$Library_preparation
sample <-  metadata_mm_discovery$sample
design = model.matrix(~group + age_fact + gender) 

## normalization + batch correction
norm_counts_mm_disco <- limma::removeBatchEffect(limma::normalizeVSN(norm.factors),
                                             batch = lib_batch,
                                             design = design) %>%
  data.frame() %>%
  rownames_to_column(var = "GeneID")
write_csv(norm_counts_mm_disco, file = here("data/norm_corrected_data/corr_norm_counts_mm_discovery.csv"))

## normalization
norm_counts_mm_disco <- limma::normalizeVSN(norm.factors) %>% 
  data.frame() %>%
  rownames_to_column(var = "GeneID")
write_csv(norm_counts_mm_disco, file = here("data/norm_corrected_data/norm_counts_mm_discovery.csv"))

## validation set
norm.factors <- DGEList(counts = filtered_counts_mm_validation, samples = metadata_mm_validation, group = metadata_mm_validation$outcome)
norm.factors <- calcNormFactors(norm.factors, method = "TMM")

group <- metadata_mm_validation$group
age <- metadata_mm_validation$age
age_fact <- metadata_mm_validation$age_fact
gender <- metadata_mm_validation$gender
lib_batch <- metadata_mm_validation$Library_preparation
sample <-  metadata_mm_validation$sample
design = model.matrix(~group + age_fact + gender) 

norm_counts_mm_valid <- limma::removeBatchEffect(limma::normalizeVSN(norm.factors), 
                                             batch = lib_batch, 
                                             design = design) %>% 
  data.frame() %>%
  rownames_to_column(var = "GeneID")
write_csv(norm_counts_mm_valid, file = here("data/norm_corrected_data/corr_norm_counts_mm_validation.csv"))

norm_counts_mm_valid <- limma::normalizeVSN(norm.factors) %>% 
  data.frame()  %>%
  rownames_to_column(var = "GeneID")

write_csv(norm_counts_mm_valid, file = here("data/norm_corrected_data/norm_counts_mm_validation.csv"))

```


```{r}
#------- HC
## discovery test
# norm.factors <- DGEList(counts = filtered_counts_hc_discovery %>% 
#                           column_to_rownames(var = "SampleID") %>% 
#                           t() %>% 
#                           data.frame(), 
#                         samples = metadata_hc_discovery, 
#                         group = metadata_hc_discovery$outcome)
# norm.factors <- calcNormFactors(norm.factors, method = "TMM")
# 
# group <- metadata_hc_discovery$group
# age <- metadata_hc_discovery$age
# age_fact <- metadata_hc_discovery$age_fact
# gender <- metadata_hc_discovery$gender
# lib_batch <- metadata_hc_discovery$Library_preparation
# sample <-  metadata_hc_discovery$sample
# design = model.matrix(~group + age_fact + gender) 
# 
# norm_counts_hc_disco <- limma::removeBatchEffect(limma::normalizeVSN(norm.factors), 
#                                              batch = lib_batch, 
#                                              design = design) %>% 
#   data.frame()
# 
# norm_counts_hc_disco <- limma::normalizeVSN(norm.factors) %>% 
#   data.frame()  %>%
#   rownames_to_column(var = "GeneID")
# 
# write_csv(norm_counts_hc_disco, file = here("data/norm_corrected_data/norm_counts_hc_discovery.csv"))
# 
# 
# ## validation set
# norm.factors <- DGEList(counts = filtered_counts_hc_validation  %>% 
#                           column_to_rownames(var = "SampleID") %>% 
#                           t() %>% 
#                           data.frame(), 
#                         samples = metadata_hc_validation,
#                         group = metadata_hc_validation$outcome)
# norm.factors <- calcNormFactors(norm.factors, method = "TMM")
# 
# group <- metadata_hc_validation$group
# age <- metadata_hc_validation$age
# age_fact <- metadata_hc_validation$age_fact
# gender <- metadata_hc_validation$gender
# lib_batch <- metadata_hc_validation$Library_preparation
# sample <-  metadata_hc_validation$sample
# design = model.matrix(~group + age_fact + gender) 
# 
# norm_counts_hc_valid <- limma::removeBatchEffect(limma::normalizeVSN(norm.factors), 
#                                              batch = lib_batch, 
#                                              design = design) %>% 
#   data.frame()
# 
# norm_counts_hc_valid <- limma::normalizeVSN(norm.factors) %>% 
#   data.frame()  %>%
#   rownames_to_column(var = "GeneID")
# 
# write_csv(norm_counts_hc_valid, file = here("data/norm_corrected_data/norm_counts_hc_validation.csv"))

```



```{r}
#------- Simulated dataset
group <- metadata_discovery$outcome

## Discovery set
norm.factors <- DGEList(counts = sim_discovery , samples = metadata_discovery , group = metadata_discovery$outcome)
norm.factors <- calcNormFactors(norm.factors, method = "TMM")

design = model.matrix(~group) 

norm_counts_sim_discovery <- limma::removeBatchEffect(limma::normalizeVSN(norm.factors), 
                                             design = design) %>% 
  data.frame()

norm_counts_sim_discovery <- limma::normalizeVSN(norm.factors) %>% 
  data.frame()  %>%
  rownames_to_column(var = "GeneID")

write_csv(norm_counts_sim_discovery, file = here("data/norm_corrected_data/norm_counts_sim_rep_discovery.csv"))

## Validation set
group = metadata_validation$outcome
norm.factors <- DGEList(counts = sim_validation, samples = metadata_validation , group = metadata_validation$outcome)
norm.factors <- calcNormFactors(norm.factors, method = "TMM")

design = model.matrix(~group) 

## Normalization + batch correction
# norm_counts_sim_validation <- limma::removeBatchEffect(limma::normalizeVSN(norm.factors), 
#                                              design = design) %>% 
#   data.frame()

## Normalization 
norm_counts_sim_validation <- limma::normalizeVSN(norm.factors) %>% 
  data.frame()  %>%
  rownames_to_column(var = "GeneID")

write_csv(norm_counts_sim_validation, file = here("data/norm_corrected_data/norm_counts_sim_validation.csv"))
write_csv(norm_counts_sim_validation, file = here("data/norm_corrected_data/corr_norm_counts_sim_validation.csv"))
```

